%!TEX root = ../template.tex
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% acknowledgements.tex
%% NOVA thesis document file
%%
%% Text with acknowledgements
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\typeout{NT FILE acknowledgements.tex}%

\begin{ntacknowledgements}

I would like to express my sincere gratitude to all those who have supported me throughout my Master's journey. This dissertation would not have been possible without their guidance, encouragement, and patience.

% --- Academic Acknowledgements ---

First and foremost, I extend my deepest gratitude to my advisor, \textbf{Professor Luís Filipe dos Santos Gomes}. I am sincerely thankful for the opportunity and for his constant availability to guide me through this entire process. A special thanks for the challenge he proposed, which allowed me to deepen my knowledge in the fields of digital systems and telecommunications, areas I am passionate about and in which I currently work. His guidance was fundamental to this project.

% --- Personal Acknowledgements ---

On a personal note, I could not have reached this point without the support of my family and friends.

To my family: thank you for making it possible for me to study and for being there during the most difficult moments. Your support gave me strength when I needed it the most.

Finally, a very special thanks to my girlfriend, \textbf{Cátia}. Thank you for always believing in me, for motivating me to push forward, and for reminding me not to give up. Your presence was essential throughout this journey.

To you all, thank you.

\end{ntacknowledgements}